---
title: "cyclistic_2023"
author: "Nurlybek Zhumabek"
date: "2024-11-28"
output: html_document
---

##Introduction

This is my first capstone project from [Google Data Analytics Professional Certificate](https://www.coursera.org/professional-certificates/google-data-analytics) course.


##Purpose
The goal of this project is to study how annual members and casual riders of a Cyclistic, bike-share company differ, and why casual riders would buy annual membership, and how digital media could affect marketing tactics. The project aims to design marketing strategies for converting casual riders to annual members. 


##Scenario

You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.


**For this project following data analysis steps will be followed:**
- Ask
- Prepare
- Process
- Analyze
- Share
- Act


##Ask

**Three questions will guide the future marketing program:**

1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

The director of marketing and your manager Lily Moreno has assigned you the first question to answer: How do annual members and casual riders use Cyclistic bikes differently?

**Key tasks**

- Identify the business task
    - The main objective is to build the best marketing strategies to turn casual bike riders into annual members by analyzing how the 'Casual' and 'Annual' customers use Cyclistic bike share differently.  

- Consider key stakeholders
    - Cyclistic executive team, Director of Marketing (Lily Moreno), Marketing Analytics team.

**Deliverable**

- A clear statement of business task
  - The business problem is how we can maximize the number of annual memberships of the Cyclistic bike share company as defined by manager, Lily moreno. I was assigned to answer the question that states how annual members and casual riders use Cyclistic bikes differently, by a marketing analytics team. My findings regarding the following question will help to drive the decision making process of the executive team for planning for implementation.


##Prepare

**Key tasks**

- Download data and store it appropriately.
  - Data had been downloaded from link provided a and stored securely on my PC. Datasets are available [here.](https://divvy-tripdata.s3.amazonaws.com/index.html) 

- Identify how it’s organized.
  - The dataset is organized by time, starting from recent to oldest data available. Each data is given as a .csv format file.

- Sort and filter the data.
  - For this analysis I'm going to use last 12 months data of the year 2023, as it's the more current period to the business task.

- Determine the credibility of the data.
  - The current dataset is recognized as credible (ROCCC: Reliable, Original, Comprehensive, Current and Cited).

**Deliverable**

- A description of all data sources used
  - I used Cyclistic's historical trip data to analyze and identify trends. The data has been made available by Motivate International Inc. under this [license](https://divvybikes.com/data-license-agreement). 


##Process

For the analysis I will be using R programming language and RStudio IDE for it's easy statistical analysis tools and data visualizations.

**Key tasks**

- Check the data for errors.
- Choose your tools.
- Transform the data so you can work with it effectively.
- Document the cleaning process.

**Deliverable**

- Documentation of any cleaning or manipulation of data
  - Please see following code chunks below


```{r installing packages}
## Open Rstudio by updated R version and install all relevant packages

library(tidyverse)
library(lubridate)
library(geosphere)
library(janitor)
library(ggmap)
```

```{r import dataset}
## Import and download dataset available from the link provided

jan23 <- read_csv("D:/Downloads/202301-divvy-tripdata_Jan.csv")
feb23 <- read_csv("D:/Downloads/202302-divvy-tripdata_Feb.csv")
mar23 <- read_csv("D:/Downloads/202303-divvy-tripdata_Mar.csv")
apr23 <- read_csv("D:/Downloads/202304-divvy-tripdata_Apr.csv")
may23 <- read_csv("D:/Downloads/202305-divvy-tripdata_May.csv")
jun23 <- read_csv("D:/Downloads/202306-divvy-tripdata_Jun.csv")
jul23 <- read_csv("D:/Downloads/202307-divvy-tripdata_Jul.csv")
aug23 <- read_csv("D:/Downloads/202308-divvy-tripdata_Aug.csv")
sep23 <- read_csv("D:/Downloads/202309-divvy-tripdata_Sep.csv")
oct23 <- read_csv("D:/Downloads/202310-divvy-tripdata_Oct.csv")
nov23 <- read_csv("D:/Downloads/202311-divvy-tripdata_Nov.csv")
dec23 <- read_csv("D:/Downloads/202312-divvy-tripdata_Dec.csv")
```

```{r check data}
## Check data for consistency

colnames(jan23)
colnames(feb23)
colnames(mar23)
colnames(apr23)
colnames(may23)
colnames(jun23)
colnames(jul23)
colnames(aug23)
colnames(sep23)
colnames(oct23)
colnames(nov23)
colnames(dec23)

## Check data stuctures
str(jan23)
str(feb23)
str(mar23)
str(apr23)
str(may23)
str(jun23)
str(jul23)
str(aug23)
str(sep23)
str(oct23)
str(nov23)
str(dec23)
```

```{r merging all dataset}
## Merging individual monthly datasets into one data frame

tripdata23 <- bind_rows(jan23,feb23,mar23,apr23,may23,jun23,jul23,aug23,sep23,oct23,nov23,dec23)
```

```{r recheck merged data frame}
## Check merged data frame

colnames(tripdata23)
head(tripdata23)
str(tripdata23)
summary(tripdata23)
```

```{r adding new columns}
## Adding date, month, year, day of week columns

tripdata23 <- tripdata23 %>% 
mutate(year=format(as.Date(started_at),"%Y")) %>% # produce year 
mutate(month=format(as.Date(started_at),"%B")) %>% # produce month 
mutate(date=format(as.Date(started_at),"%d")) %>% # produce date 
mutate(day_of_week=format(as.Date(started_at),"%A")) %>% # produce day of week       
mutate(ride_length=difftime(ended_at,started_at)) %>% 
mutate(start_time=strftime(started_at, "%H"))

## Converting “ride_length” to numeric data type for calculation

tripdata23 <- tripdata23 %>% mutate(ride_length=as.numeric(ride_length))

is.numeric(tripdata23$ride_length) # recheck
```

```{r adding ride distance}
## Adding ride distance in km

tripdata23$ride_distance <- distGeo(matrix(c(tripdata23$start_lng,tripdata23$start_lat),ncol = 2),matrix(c(tripdata23$end_lng,tripdata23$end_lat),ncol = 2))

tripdata23$ride_distance <- tripdata23$ride_distance/1000 #distance in km
```

```{r cleaning procedures}
## Data cleaning procedures

colSums(is.na(tripdata23)) # check for na values

tripdata23_cln1 <- tripdata23[!is.na(tripdata23$end_lat),] #remove na values in “eng_lat”

colSums(is.na(tripdata23_cln1)) # created new cleaned data frame and rechecked


## Checking for blanks written as "". Because I noticed them by checking for consistency above.

sum(tripdata23_cln1$start_station_name == '')
sum(tripdata23_cln1$start_station_id == '')
sum(tripdata23_cln1$end_station_id == '')
sum(tripdata23_cln1$end_station_name == '')


## Delete those blanks ““ by creating another new data cleaned frames

tripdata23_cln2 <- tripdata23_cln1[trimws(tripdata23_cln1$start_station_name) !='',]
tripdata23_cln3 <- tripdata23_cln2[trimws(tripdata23_cln2$start_station_id) !='',]
tripdata23_cln4 <- tripdata23_cln3[trimws(tripdata23_cln3$end_station_name) !='',]
tripdata23_cln5 <- tripdata23_cln4[trimws(tripdata23_cln4$end_station_id) !='',]

## Remove those data frames and keep the last cleaned version of data frame

rm(tripdata23_cln4)
rm(tripdata23_cln3)
rm(tripdata23_cln2)
rm(tripdata23_cln1)
```


##Analyze

**Key tasks**

- Aggregate your data so it’s useful and accessible.
- Organize and format your data.
- Perform calculations.
- Identify trends and relationships.

**Deliverable**

- A summary of analysis
  - Please see following code chunks below and read provided summary for each step
  
Conduct descriptive analysis

```{r determining mean,max,min,median}
# Determining mean, max, min, median

tripdata23_cln5 %>% 
  summarise(average_ride_length=mean(ride_length),median_length=median(ride_length), max_ride_length=max(ride_length),min_ride_length=min(ride_length))
```

- This above data is about 'ride_length' depending on the whole year 2023. Minimum ride length (min_ride_length) and Maximum ride length (max_ride_length) has absurd values, due to the lack of scope it not possible to find out the problem behind it, but it need to be analyzed further.


Compare members and casual users
```{r compare members and casuals}
## Members vs casual riders difference depending on total rides taken

tripdata23_cln5 %>% group_by(member_casual) %>% summarise(ride_count=length(ride_id),ride_percentage=(length(ride_id)/nrow(tripdata23_cln5))*100)

## Casual vs member riders’ distribution (visualization)

ggplot(tripdata23_cln5, aes(x=member_casual,fill = member_casual)) + geom_bar() + labs(x="Casuals vs Members",y="Number of Rides",title = "Casuals vs Members distribution")
```

- We can see on the Casuals vs Members distribution chart, members possesing ~65%, and casual riders have ~35% of the dataset. So it is clearly visible that in the whole year 2023 members used ride share ~30% more than casual riders.


Comparison between Members Causal riders depending on ride length (mean, median, minimum, maximum)
```{r comparison by mean,median,min,max of casuals and members}
tripdata23_cln5 %>% group_by(member_casual) %>% summarise(average_ride_length=mean(ride_length),median_length=median(ride_length),max_ride_length=max(ride_length),min_ride_length=min(ride_length))
```

- From the above table we can conclude that casual riders took bike for longer rides than members, as the average trip duration / average ride length of member riders is lower than the average trip duration / average ride length of casual riders.


See total rides and average ride time by each day for members vs casual riders
```{r fix days of week order}
## lets fix days of week order

tripdata23_cln5$day_of_week <- ordered(tripdata23_cln5$day_of_week,levels=c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))

tripdata23_cln5 %>% 
group_by(member_casual,day_of_week) %>% 
summarise(number_of_rides = n(),average_ride_length=mean(ride_length),.groups = "drop") %>% arrange(member_casual,day_of_week)
```


Visualize total rides data by type and day of week
```{r total rides by type for week days}
tripdata23_cln5 %>% 
group_by(member_casual,day_of_week) %>% 
summarise(number_of_rides=n(),.groups = "drop") %>% 
arrange(member_casual,day_of_week) %>% ggplot(aes(x=day_of_week,y=number_of_rides,fill=member_casual)) + 
labs(title = "Total rides by Members and Casual riders Vs. Day of the week") + geom_col(width = 0.5,position = position_dodge(width = 0.5)) + 
scale_y_continuous(labels = function(x) format(x,scientific=FALSE))
```

Visualize average ride time data by type and day of week
```{r average ride time}
tripdata23_cln5 %>% 
group_by(member_casual,day_of_week) %>% summarise(average_ride_length=mean(ride_length),.groups = "drop") %>% ggplot(aes(x=day_of_week,y=average_ride_length,fill=member_casual)) + 
geom_col(width = 0.5,position = position_dodge(width = 0.5)) + 
labs(title = "Average ride time by Members and Casual riders Vs. Day of the week")
```

- From the first chart above members took consistent trips throughout the week, but there is less rides in Saturday, Sunday. It has more like a bell-shaped curve. For casual riders the most taken rides are in weekends, starting rise in Friday followed by Saturday and Sunday. But here it clearly obvious and shown that member rides had been rocked up in recent years and casuals decresed abruptly. 

- The average ride length (time) for members are less than that of casual riders. Also it can be seen that weekend average ride (time) length is much higher for casual riders along with total rides. So both of this facts can be correlated for casual riders. For members average ride lengh is about the same throughout the week (<1000 sec).



See total rides and average ride time by each month for members vs casual riders
```{r fix month of year order}
## Lets fix months of year order

tripdata23_cln5$month <- ordered(tripdata23_cln5$month,levels=c("January","February","March","April","May","June","July","August","September","October","November","December")) 

tripdata23_cln5 %>% 
group_by(member_casual,month) %>% summarise(number_of_rides=n(),average_ride_length=mean(ride_length),.groups = "drop") %>% arrange(member_casual,month)
```


Visualize total rides data by type and month
```{r total rides by type and month}
tripdata23_cln5 %>% 
group_by(member_casual,month) %>% 
summarise(number_of_rides=n(),.groups = "drop") %>% 
arrange(member_casual,month) %>% ggplot(aes(x=month,y=number_of_rides,fill=member_casual)) + labs(title = "Total rides by Members and Casual riders Vs. Month",x="Month",y="Number of Rides") + 
theme(axis.title.x = element_text(angle = 45)) + 
geom_col(width = 0.5,position = position_dodge(width = 0.5)) + 
scale_y_continuous(labels = function(x) format(x,scientific=FALSE))
```

Visualize average ride time data by type and month
```{r average ride time by months}
tripdata23_cln5 %>% 
group_by(member_casual,month) %>% 
summarise(average_ride_length=mean(ride_length),.groups = "drop") %>% ggplot(aes(x=month,y=average_ride_length,fill = member_casual))+geom_col(width = 0.5,position = position_dodge(width = 0.5)) + 
labs(title = "Average ride length by Members and Casual riders Vs. Month") + theme(axis.title.x = element_text(angle = 30))
```

- The months June, July, August and September are the most busy time of the year among both members and casual riders. It is possible due to winter there is a significant drop in total rides in the months of November, December, January and February for both type of customers. But we can see that member's total rides are higher than casual riders throughout the year. Both members and casual has bell-shaped figure.

- Average ride length of members is about the same <7000 secs throughout the year. While casual riders average ride length is between 700 - 1500 secs throughout the year.But we can see that casual's total rides are higher than member riders throughout the year. Members have stable level oof ride time. Casuals have bell-shaped figure. 


Comparison between Members and Casual riders depending on ride distance
```{r ride distance comparison}
tripdata23_cln5 %>% group_by(member_casual) %>% summarise(average_ride_distance=mean(ride_distance)) %>% 
ggplot() + geom_col(mapping = aes(x=member_casual,y=average_ride_distance,fill = member_casual), show.legend = FALSE) + 
labs(title="Mean travel distance by Members and Casual riders",x="Member and Casual riders",y="Average distance In Km")
```

- From the above chart we can see that both riders travel about the same average distance. This similarity could be possible due to that member take (same ride time) rides throughout the week, but casual riders took rides mostly in weekends with higher ride time.


Analysis and visualization on cyclistic's bike demand by hour in a day
```{r hour in a day}
tripdata23_cln5 %>% ggplot(aes(start_time,fill=member_casual)) + 
labs(x="Hour of the day",title = "Cyclistic's Bike demand by hour in a day") + geom_bar()
```

- From the above chart we can see both more members and more casuals between 7am and 09am and between 3pm and 7pm. Both of them have similar figures in shape, but number casuals a higher than number of members. Also there is bigger volume rise in the morning for both type of riders. This information needs to be checked on day basis. Also at the end of noon and beginning of evening we can see higher count of rides for both type riders.


Analysis and visualization on cyclistic's bike demand per hour by day of the week
```{r hours by day of the week}
tripdata23_cln5 %>% ggplot(aes(start_time,fill = member_casual)) + geom_bar() + labs(x="Hour of the day",title = "Cyclistic's bike demand per hour by day of the week") + facet_wrap(~ day_of_week)
```

- There is a lot of diferrence between the weekdays and weekends. There is a big increase of volume in the weekdays between 7am to 10am and another volume increase from 5pm to 7pm. We can hypothesize that members use the bikes as daily routine like going to work (same behaviour throughout the weekdays) and go back from work (5pm - 7pm). Weekends are completely different for members and casual riders, Friday, Saturday and Sunday there is huge peak in volume for casual riders, from this we can hypothesize that casual riders mostly use bike share for leisure activity in the weekends.


Analysis and visualization of Rideable type Vs. total rides by Members and casual riders
```{r rideable type}
tripdata23_cln5 %>% group_by(rideable_type) %>% summarise(count = length(ride_id))

ggplot(tripdata23_cln5, aes(x=rideable_type,fill = member_casual)) +
labs(x="Rideable type", title = "Rideable type Vs. total riders by Members and casual riders") + geom_bar()
```

- From the above viz we can see that members mostly use classic bikes, followed by electric bikes Docked bikes mostly used by casual riders. Electric bikes are more favored by members.


Now analyze and visualize the dataset on coordinate basis
```{r coordinates}
## Lets check the coordinates data of the rides.
## Adding a new data frame only for the most popular routes >200 rides

coordinates_df <- tripdata23_cln5 %>% filter(start_lng !=end_lng & start_lat !=end_lat) %>% group_by(start_lng, start_lat, end_lng, end_lat, member_casual, rideable_type) %>% summarise(total_rides = n(),.groups = "drop") %>% filter(total_rides>200)

# now lets create two different data frames depending on rider type (member_casual)

casual_riders <- coordinates_df %>% filter(member_casual=="casual")
member_riders <- coordinates_df %>% filter(member_casual=="member")
```

Lets setup ggmap and store map of Chicago (bbox, stamen map)
```{r stamen map}
chicago <- c(left = -87.700424, bottom = 41.790769, right = -87.554855, top = 41.990119)
```

Visualization on the map
```{r API key}
## Registered on website of Stadiamap and received API key:

register_stadiamaps(key = "d031d9b2-74eb-4a34-9889-ec5d366984ab",write = FALSE)
stadiamaps_key()
has_stadiamaps_key()

chicago_map <- get_stadiamap(bbox = chicago, zoom = 12, maptype = "stamen_terrain")
```

```{r viz on both types of riders on map}
## Maps on casual riders

ggmap(chicago_map,darken = c(0.1,"white")) + geom_point(casual_riders,mapping = aes(x=start_lng,y=start_lat,color=rideable_type),size=2) + coord_fixed(0.8) + labs(title = "Most used routes by Casual riders",x=NULL,y=NULL) + theme(legend.position = "none")

## Maps on member riders

ggmap(chicago_map,darken = c(0.1,"white")) + geom_point(member_riders,mapping = aes(x=start_lng,y=start_lat,color=rideable_type),size=2) + coord_fixed(0.8) + labs(title = "Most used routes by Member riders",x=NULL,y=NULL) + theme(legend.position = "none")
```

Coordinate system already present. Adding new coordinate system, which will replace the existing one.
Warning message:
Removed 9 rows containing missing values or values outside the scale range (`geom_point()`).

Coordinate system already present. Adding new coordinate system, which will replace the
existing one.
Warning message:
Removed 46 rows containing missing values or values outside the scale range
(`geom_point()`).

- We can clearly see the casual rides are mostly located around the center of the town (or the bay area), with all their trips located around that area points towards their bike usage pattern, which is for leisure, probably tourist or sightseeing related rides.

- Members are mostly use bike all over the city including main city area and outside main center. This can be hypothesize as they travel for work purpose.



##Share

**Key tasks**

- Determine the best way to share your findings.
- Create effective data visualizations.
- Present your findings.
- Ensure your work is accessible.

**Deliverable**

- Supporting visualizations and key findings


**Main insights and finding conclusions**

- Members holds the biggest proportion of the total rides, ~30% bigger than casuals riders.
- In all months we have more members than casual riders.
- For casual riders the biggest volume of data is on the the weekend.
- There is a bigger volume of bikers in the afternoon.

This could be possible that member use bikes for work purpose, this information can be backed by their bike usage in colder months, where there is significant drop in casual members in those months.


**Now for how members differs from casuals:**

- Members have the bigger volume of data, except on saturday and sunday. On the weekend, casuals riders have the most data points.
- Casuals riders have more ride length (ride duration) than members. Average ride time of member are mostly same slight increase in end of week.
- We have more members during the morning, mainly between 7am and 9am. And more casuals between 3pm and 2am. Actually this figure is applicable for both types of riders.
- Members have a bigger preference for classic bikes, followed by electric bike.
- Members have a more fixed use for bikes for routine activities. Where as casual rider's usage is different, mostly all activiy in the weekend.
- Casual member spend time near the center of the city or the bay area, where as member are scattered throughout the city.


##Act

Act phase will be done by the Cyclistic's executive team, Director of Marketing (Lily Moreno), Marketing Analytics team on the basis of my analysis. (Data-driven decision making)

**Deliverable**
 
Top three recommendations based on your analysis

- Offer a weekend-only membership at a different price point than the full annual membership.
- Coupons and discounts could be handed out along with the annual subscription / weekend-only membership for the usage of electric bikes targeting casual riders. Also increasing the number of electric bike while reducing classic bikes if electric bike costs more for the pass, this can be beneficial for the company. (As electric bike are already in trend and usage is good as per member and ride type data.
- Create marketing campaigns which can be sent via email, or advertisement in the docking stations explaining why annual member is beneficial. Campaigns should be placed at the peak months of the year.

**Note : All ride ids are unique so we cannot conclude if the same rider taken several rides. More rider data needed for further analysis**


Additonal data that could expand scope of analysis

- Pricing details for members and casual riders - Based on this data, we might be to optimize cost structure for casual riders or provide discounts without affecting the profit margin.
- Address/ neighborhood details of members to investigate if there are any location specific parameters that encourage membership.
- Way to determine a recurring bike user using payment information or any personal identification.
